Modelling BCG vaccination in the UK

### 2019-08-02 rmds/07-direct-effect-bcg-change.Rmd
updated links to preprints
direct effects hannah comments

I combined notification data from the Enhanced TB Surveillance (ETS) system, with demographic data from the Labour Force Survey to construct retrospective cohorts of individuals in England relevant to both the universal, and targeted vaccination programmes between Jan 1, 2000 and Dec 31, 2010. For each cohort, I estimated incidence rates over a 5 year follow-up period and used Poisson and Negative Binomial regression models in order to estimate the impact of the change in policy on TB. This work was supervised by Hannah Christensen and Ellen Brooks-Pollock. Nicky Weltton provided guidance on the statistical methods used.

@@

In 2005 England changed its Bacillus Calmette–Guérin (BCG) vaccination policy against TB (TB) from a universal programme aimed at 13 and 14 year olds to a targeted programme aimed at high-risk neonates (see Chapter \@ref(background)). High risk babies are identified by local TB incidence and by the parents’ and grandparents’ country of origin. The change in policy was motivated by evidence of reduced TB transmission,[@PHE2016a; @Fine2005a; @Teo2006] and high effectiveness of the BCG vaccine in children,[@Rodrigues1993; @Colditz1994; @Roy2014] and variable effectiveness in adults.[Mangtani2014; @Zwerling2011] Little work has been done to evaluate the impact of this change in vaccination policy.

@@

I obtained yearly population estimates from the April to June Labour Force Survey (LFS) for 2000-2015. The LFS is a study of the employment circumstances of the UK population, and provides the official measures of employment and unemployment in the UK (see Chapter \@ref(data)). Reporting practices have changed with time so the appropriate variables for age, country of origin, country of birth, and survey weight were extracted from each yearly extract, standardised, and combined into a single data-set.

@@

I calculated Incidence Rate Ratios (IRRs) for the change in incidence rates associated with the change in BCG vaccination policy (modelled as a binary breakpoint at the start of 2005) for both the UK born and non-UK born populations that were relevant to the universal programme, and for the targeted programme using a range of models. I considered the following covariates: age,[@PHE2016a; @Zwerling2011] incidence rates in both the UK born and non-UK born who were not in the age group of interest,[@PHE2016a] and year of study entry (as a random intercept). I first investigated a univariable Poisson model, followed by combinations of covariates (Table \@ref(tab:07-model-spec)). I also investigated a Negative Binomial model adjusting for the same covariates as in the best fitting Poisson model. The models were estimated with a Bayesian approach using Markov Chain Monte Carlo (MCMC), with default weakly informative priors (see  Section \@ref(prior-choice)). Model fit, penalised by model complexity, was assessed using the leave one out cross validation information criterion (LOOIC) and its standard error.[@Vehtari2016] Models were ranked by goodness of fit, using their LOOIC, with a smaller LOOIC indicating a better fit to the data after adjusting for the complexity of the model. No formal threshold for a change in the LOOIC was used, with changes in the LOOIC being evaluated in the context of their standard error. The inclusion of the change in policy in the best fitting model was tested by refitting the model excluding the change in policy and estimating the improvment in the LOOIC. Once the best fitting model had been identified I estimated the number of cases prevented, from 2005 until 2015, for each vaccination programme in the study population relevant to that programme (see Section \@ref(magnitude-estimation)).

@@

R `3.5.0` was used for all analysis.[@R] Reproducibility was ensured by using R package infrastructure (https://github.com/seabbs/DirectEffBCGPolicyChange). Missing data imputation using a GBM was implemented using the `h2o` package (see Section \@ref(imp-uk-birth-status)).[@h2o2018] Incidence rates, with 95% confidence intervals, were calculated using the `epiR` package (see Chapter \@ref(data))^[An R package is available containing the required code DOI: [10.5281/zenodo.2551554](https://zenodo.org/badge/latestdoi/93072437)].[@EpiR] The `brms` package,[@Burkner] and STAN,[@StanDevelopmentTeam2016] was used to perform MCMC. Models were run until convergence (4 chains with a burn in of 10,000, and 10,000 sampled iterations each), with convergence being assessed using trace plots and the R hat diagnostic.[@StanDevelopmentTeam2016] All numeric confounders were centered and scaled by their standard deviation, and age was adjusted for using single year of age categories.

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of techniques from machine learning to improve the quality of our imputation, whilst also validating it using metrics supported by theory. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration data set. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

Where $C^i_P$ is the predicted number of cases prevented using the mean ($e$), 2.5% bound ($l$) and 97.5% bound ($u$) of the IRR estimate $I^i$. For the cohorts relevant to the targeted high-risk neonatal scheme I used a related equation, adjusting for the fact that the populations were exposed to the scheme and I therefore had to first estimate the number of cases that would have been observed had the scheme not been implemented. After simplification this results in the following equation,

@@

```{r 07-summary-universal-nonukborn, results = "asis"}
read_formated("universal-nonukborn-results-summary") %>%
  make_summary_table(
    caption.short = "Comparison of models fitted to incidence rates for the non-UK born population that were eligible to the universal vaccination programme of those at school-age (14).", 
    caption = paste0("Comparison of models fitted to incidence rates for the non-UK born population that were eligible to the universal vaccination programme of those at school-age (14). ", ModelOrderingString))
```

@@

For the UK born cohort relevant to the targeted neonatal vaccination programme the evidence of an association, across all models, was mixed and credible intervals were wide compared to models for the UK born cohort relevant to the universal school-age vaccination programme (Table \@ref(tab:07-summary-targeted-ukborn)). The best fitting model was a Poisson model which adjusted for the change in policy, age, UK born incidence rates, and year of study entry with a random effect (Table \@ref(tab:07-targeted-tab)). In this model, there was weak evidence of an association between the change in BCG policy and an decrease in incidence rates in UK born neonates, with an IRR of `r extract_inc_ef(ModelComTargetedUKBorn, ModelComTargetedUKBorn$Model[1])`. There was weak evidence to suggest that dropping the change in policy from this model improved the quality of the fit, with an improvement in the LOOIC score of `r BestFitTargetedUKBorn[[5]]$pretty_diff_looic`. This suggests that the change in policy was not an important factor for explaining incidence rates, after adjusting for covariates. Models which also adjusted for non-UK born incidence rates estimated that the change in policy was associated with no change in incidence rates in the relevant cohort of neonates.

@@

```{r 07-summary-targeted-ukborn}
read_formated("targeted-ukborn-results-summary") %>%  
  make_summary_table(
    caption.short = "Comparison of models fitted to incidence rates for the UK born population that were elgible to the targeted vaccination programme of neonates.", 
    caption = paste0("Comparison of models fitted to incidence rates for the UK born population that were elgible to the targeted vaccination programme of neonates. ", ModelOrderingString))
```

@@

I estimate that the change in vaccination policy was associated with preventing `r pretty_inline_ci(EstMagImpactBestModel[[7, 3]], note = "95%CI ")` cases from 2005 until the end of the study period in the directly impacted populations after 5 years of follow up (Table \@ref(tab:07-tab-impact)). The majority of the cases prevented were in the non-UK born, with cases increasing slightly overall in the UK born. This was due to cases increasing in the UK born at school-age, and decreasing in UK born neonates, although both these estimates had large credible intervals.

@@

It is well established that interventions against infectious diseases, such as TB, should be evaluated not only for their direct effects but also for future indirect effects via ongoing transmission. Statistical approaches such as those used in this paper are not appropriate for capturing these future indirect effects, and instead dynamic disease models should be used. In Chapter \@ref(model-development) I develop such a dynamic disease model, Chapter \@ref(model-fitting) then fits this model to the available data, and Chapter \@ref(model-results) compares the impact of continuing with the BCG school's scheme post 2005 compared to universal neonatal vaccination. In addition, this study could not evaluate the impact of the neonatal programme on the high-risk population it targets, due to a lack of reliable data. This is a limitation of all of the analyses presented in this thesis. Improved coverage data for the BCG programme is required to more fully evaluate its ongoing impact.

>>>>

I combined notification data from the Enhanced TB Surveillance (ETS) system, with demographic data from the Labour Force Survey (LFS) to construct retrospective cohorts of individuals in England relevant to both the universal, and targeted vaccination programmes between Jan 1, 2000 and Dec 31, 2010. For each cohort, I estimated incidence rates over a 5 year follow-up period and used Poisson and Negative Binomial regression models in order to estimate the impact of the change in policy on TB. This work was adapted from a preprint^[Preprint: https://doi.org/10.1101/567511] supervised by Hannah Christensen and Ellen Brooks-Pollock. Nicky Welton provided guidance on the statistical methods used.

@@

In 2005 England changed its Bacillus Calmette–Guérin (BCG) vaccination policy against tuberculosis (TB) from a universal programme aimed at 13 and 14 year olds to a targeted programme aimed at high-risk neonates (see Chapter \@ref(background)). High risk babies are identified by local TB incidence and by the parents’ and grandparents’ country of origin. The change in policy was motivated by evidence of reduced TB transmission,[@PHE2016a; @Fine2005a; @Teo2006] and high effectiveness of the BCG vaccine in children,[@Rodrigues1993; @Colditz1994; @Roy2014] and variable effectiveness in adults.[Mangtani2014; @Zwerling2011] Little work has been done to evaluate the impact of this change in vaccination policy.

@@

I obtained yearly population estimates from the April to June LFS for 2000-2015. The LFS is a study of the employment circumstances of the UK population, and provides the official measures of employment and unemployment in the UK (see Chapter \@ref(data)). Reporting practices have changed with time so the appropriate variables for age, country of origin, country of birth, and survey weight were extracted from each yearly extract, standardised, and combined into a single data-set (see Section \@ref(demographic-data)).

@@

I calculated Incidence Rate Ratios (IRRs) for the change in incidence rates associated with the change in BCG vaccination policy (modelled as a binary breakpoint at the start of 2005) for both the UK born and non-UK born populations that were relevant to the universal programme, and for the targeted programme using a series of increasingly complex models. I considered the following covariates: age,[@PHE2016a; @Zwerling2011] incidence rates in both the UK born and non-UK born who were not in the age group of interest,[@PHE2016a] and year of study entry (as a random intercept). I first investigated a univariable Poisson model, followed by combinations of covariates (Table \@ref(tab:07-model-spec)). I also investigated a Negative Binomial model adjusting for the same covariates as in the best fitting Poisson model. The models were estimated with a Bayesian approach using Markov Chain Monte Carlo (MCMC), with default weakly informative priors (see  Section \@ref(prior-choice)). Model fit, penalised by model complexity, was assessed using the leave one out cross validation information criterion (LOOIC) and its standard error.[@Vehtari2016] Models were ranked by goodness of fit, using their LOOIC, with a smaller LOOIC indicating a better fit to the data after adjusting for the complexity of the model. No formal threshold for a change in the LOOIC was used, with changes in the LOOIC being evaluated in the context of their standard error. The inclusion of the change in policy in the best fitting model was tested by refitting the model excluding the change in policy and estimating the improvment in the LOOIC. Once the best fitting model had been identified I estimated the number of cases prevented, from 2005 until 2015, for each vaccination programme in the study population relevant to that programme (see Section \@ref(magnitude-estimation)).

@@

R `3.5.0` was used for all analysis.[@R] Reproducibility was ensured by using `R` package infrastructure^[Analysis R package DOI: [10.5281/zenodo.2551554](https://zenodo.org/badge/latestdoi/93072437)]. Missing data imputation using a GBM was implemented using the `h2o` package (see Section \@ref(imp-uk-birth-status)).[@h2o2018] Incidence rates, with 95% confidence intervals, were calculated using the `epiR` package (see Chapter \@ref(data)).[@EpiR] The `brms` package,[@Burkner] and STAN,[@StanDevelopmentTeam2016] was used to perform MCMC. Models were run until convergence (4 chains with a burn in of 10,000, and 10,000 sampled iterations each), with convergence being assessed using trace plots and the R hat diagnostic.[@StanDevelopmentTeam2016] All numeric confounders were centered and scaled by their standard deviation, and age was adjusted for using single year of age categories.

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of more complex, high-performing models compared to those usually used for imputation, whilst also allowing the results to be validated it using predictive modelling performance metrics. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration data set. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

Where $C^i_P$ is the predicted number of cases prevented using the mean ($e$), 2.5% bound ($l$) and 97.5% bound ($u$) of the IRR estimate ($I^i$). For the cohorts relevant to the targeted high-risk neonatal scheme I used a related equation, adjusting for the fact that the populations were exposed to the scheme and I therefore had to first estimate the number of cases that would have been observed had the scheme not been implemented. After simplification this results in the following equation,

@@

```{r 07-summary-universal-nonukborn, results = "asis"}
read_formated("universal-nonukborn-results-summary") %>%
  make_summary_table(
    caption.short = "Comparison of models fitted to incidence rates for the non-UK born population that were eligible for the universal vaccination programme of those at school-age (14).", 
    caption = paste0("Comparison of models fitted to incidence rates for the non-UK born population that were eligible for the universal vaccination programme of those at school-age (14). ", ModelOrderingString))
```

@@

For the UK born cohort relevant to the targeted neonatal vaccination programme the evidence of an association between the change in policy and TB incidence was mixed across all models and credible intervals were wide compared to models for the UK born cohort relevant to the universal school-age vaccination programme (Table \@ref(tab:07-summary-targeted-ukborn)). The best fitting model was a Poisson model which adjusted for the change in policy, age, UK born incidence rates, and year of study entry with a random effect (Table \@ref(tab:07-targeted-tab)). In this model, there was weak evidence of an association between the change in BCG policy and an decrease in incidence rates in UK born neonates, with an IRR of `r extract_inc_ef(ModelComTargetedUKBorn, ModelComTargetedUKBorn$Model[1])`. There was weak evidence to suggest that dropping the change in policy from this model improved the quality of the fit, with an improvement in the LOOIC score of `r BestFitTargetedUKBorn[[5]]$pretty_diff_looic`. This suggests that the change in policy was not an important factor for explaining incidence rates, after adjusting for covariates. Models which also adjusted for non-UK born incidence rates estimated that the change in policy was associated with no change in incidence rates in the relevant cohort of neonates.

@@

```{r 07-summary-targeted-ukborn}
read_formated("targeted-ukborn-results-summary") %>%  
  make_summary_table(
    caption.short = "Comparison of models fitted to incidence rates for the UK born population that were eligible for the targeted vaccination programme of neonates.", 
    caption = paste0("Comparison of models fitted to incidence rates for the UK born population that were eligible for the targeted vaccination programme of neonates. ", ModelOrderingString))
```

@@

I estimate that the change in vaccination policy was associated with preventing `r pretty_inline_ci(EstMagImpactBestModel[[7, 3]], note = "95%CI ")` cases from 2005 until the end of the study period (2015) in the directly impacted populations with 5 years of follow up (Table \@ref(tab:07-tab-impact)). The majority of the cases prevented were in the non-UK born, with cases increasing slightly overall in the UK born. This was due to cases increasing in the UK born at school-age, and decreasing in UK born neonates, although both these estimates had large credible intervals.

@@

It is well established that interventions against infectious diseases, such as TB, should be evaluated not only for their direct effects but also for future indirect effects via ongoing transmission. Statistical approaches such as those used in this paper are not appropriate for capturing these future indirect effects, and instead dynamic disease models should be used. In Chapter \@ref(model-development) I develop such a dynamic disease model, Chapter \@ref(model-fitting) then fits this model to the available data, and Chapter \@ref(model-results) compares the impact of continuing with the BCG school's scheme post 2005 compared to universal neonatal vaccination. In addition, this study could not evaluate the impact of the neonatal programme on the high-risk population it targets, due to a lack of reliable data. Improved coverage data for the BCG programme is required to more fully evaluate its ongoing impact.

### 2019-08-02 rmds/98-discussion.Rmd
direct effects hannah comments

In Chapter \@ref(data) I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out) I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland) I explored some of the evidence that was used by policy makers to assess the impact of the change in policy. I found that the previous approach had drastically underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending the scheme was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff) where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations. However, in this chapter I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Although I was unable to rule out an unrelated policy change. Finally in Chapter \@ref(model-results) I found that the BCG schools' scheme was projected to reduce overall TB incidence compared to both neonatal and no vaccination over a range of time horizons. However, neonatal vaccination reduced incidence in children compared to any other scenario although it had little impact in any other age group. No vaccination programme evaluated had an impact on incidence in older adults. The results from Chapter \@ref(model-results) are preliminary only as the model on which they were based was a very poor fit to the observed data (Chapter \@ref(model-fitting)).

@@

Public engagement has also been a primary focus of this thesis. This is closely linked with the previous aim of open and reproducible research. Effort has been taken so that all peer reviewed content is available for the wider public, with Twitter used disseminate findings. Where appropriate, interactive applications have been developed that seek to explore some of the key findings of this thesis, as well as teaching more theoretical concepts used throughout (see Chapter \@ref(introduction)). Numerous case studies have also been produced that outline these theoretical concepts using some of the open source tools developed alongside this thesis^[See my personal site: https://www.samabbott.co.uk]. These tools were themselves developed to lower the barrier of entry to infectious disease research. One of these tools, `idmodelr`, has been released to CRAN. Finally, in 2017 I spent a week at the Green Man Festival exploring the mathematics of vaccination with the general public. This made use of several simple games, as well as an interactive online tool^[Available here: https://github.com/seabbs/pebblegame].

@@

This thesis has provided new evidence concerning the 2005 change in BCG policy in England. A simulation study that was used as part of the quantitative evidence for the change in policy was recreated and updated. The results from this updated model suggested that the change in policy was likely to have a greater impact on the UK born, at school-age, than previously thought. This finding was supported by a regression modelling study on the impact on TB incidence rates from the policy change. However, this study also found that the change in policy was associated with some benefits in UK born neonates and a much larger reduction in TB incidence rates in both non-UK born neonates and those at school-age. An additional regression study looking at the possible link between BCG vaccination and improved TB outcomes found some evidence that BCG vaccination was associated with reduced all-cause mortality, with little evidence of any other benefits. This result strengthens the case for wider vaccination. Additionally, a dynamic model of TB transmission was developed to provide a more detailed tool for evaluating the impact of the change in policy. Although this model was a poor fit to the observed data there was some evidence that continuing school-age vaccination would have resulted in fewer overall TB cases but that neonatal vaccination reduced TB incidence in children compared to continuing school-age vaccination. There was also some evidence that none of the vaccination strategies evaluated impacted long term incidence rates in older adults. These findings suggest a stronger case for the use of the BCG vaccine in school-age populations; in areas with an equivalent level of TB transmission to England; and where the effectiveness of the BCG vaccine has been shown to not reduce with age. They also indicate that a future vaccine, without the reduced effectiveness observed in some geographic areas, may be more effectively targeted at those at school-age that at neonates. However, this depends on the potential duration of protection inferred by vaccination. In addition, they highlight the trade-off between neonatal and school-age vaccination with school-age vaccination reducing overall TB incidence but neonatal vaccination reducing incidence in young children where more severe outcomes are more common. The findings from this thesis may be of use to policy-makers to inform vaccine usage both in the UK and globally. As a by-product of the work conducted in this thesis several open source tools have been developed. These tools maybe used as learning resources, for public engangement, and as part of other research projects.

>>>>

In Chapter \@ref(data), I found that there was some evidence that negative outcomes were more frequent in Tuberculosis (TB) cases not BCG vaccinated than in those that were. I also found that missingness in routine surveillance sources of TB data was associated with multiple risk factors. In Chapter \@ref(beneficial-bcg-out), I found supporting evidence that BCG vaccination was associated with reduced all-cause morality with some evidence that this may have been due to reduced TB mortality. I found little evidence for any other association with TB outcomes, after adjusting for confounding. In Chapter \@ref(sutherland), I explored some of the evidence that was used by policy makers to assess the impact of the change in policy. I found that the previous approach had drastically underestimated the amount of uncertainty surrounding the effect estimates. Using newly available data, I also found that ending the scheme was projected to result in greater number of notifications in the UK born than previously thought. These findings were confirmed in Chapter \@ref(direct-eff), where I evaluated the evidence in the surveillance data that the change in policy had impacted TB incidence rates in the target populations. However, in this chapter I found that any increase in TB notifications in the UK born was likely far outweighed by reductions in the number of notifications in the non-UK born. Although I was unable to rule out an unrelated policy change. Finally in Chapter \@ref(model-results), I found that the BCG schools' scheme was projected to reduce overall TB incidence compared to both neonatal and no vaccination over a range of time horizons. However, neonatal vaccination reduced incidence in children compared to any other scenario although it had little impact in any other age group. No vaccination programme evaluated had an impact on incidence in older adults. The results from Chapter \@ref(model-results) are preliminary only as the model on which they were based was a very poor fit to the observed data (Chapter \@ref(model-fitting)).

@@

Public engagement has been a constant theme throughout my doctoral work. This is closely linked with the previous aim of open and reproducible research. Effort has been taken so that all peer reviewed content is available for the wider public, with Twitter used disseminate findings. Where appropriate, interactive applications have been developed that seek to explore some of the key findings of this thesis, as well as teaching more theoretical concepts used throughout (see Chapter \@ref(introduction)). Numerous case studies have also been produced that outline these theoretical concepts using some of the open source tools developed alongside this thesis^[See my personal site: https://www.samabbott.co.uk]. These tools were themselves developed to lower the barrier of entry to infectious disease research. One of these tools, `idmodelr`, has been released to CRAN. Finally, in 2017 I spent a week at the Green Man Festival exploring the mathematics of vaccination with the general public. This made use of several simple games, as well as an interactive online tool^[Available here: https://github.com/seabbs/pebblegame].

@@

This thesis has provided new evidence regarding the use of BCG vaccination in England. A simulation study that was used as part of the quantitative evidence for the change in policy was recreated and updated. The results from this updated model suggested that the change in policy was likely to have a greater impact on the UK born, at school-age, than previously thought. This finding was supported by a regression modelling study on the impact on TB incidence rates from the policy change. However, this study also found that the change in policy was associated with some benefits in UK born neonates and a much larger reduction in TB incidence rates in both non-UK born neonates and those at school-age. An additional regression study looking at the possible link between BCG vaccination and improved TB outcomes found some evidence that BCG vaccination was associated with reduced all-cause mortality, with little evidence of any other benefits. This result strengthens the case for wider vaccination. Additionally, a dynamic model of TB transmission was developed to provide a more detailed tool for evaluating the impact of the change in policy. Although this model was a poor fit to the observed data there was some evidence that continuing school-age vaccination would have resulted in fewer overall TB cases but that neonatal vaccination reduced TB incidence in children compared to continuing school-age vaccination. There was also some evidence that none of the vaccination strategies evaluated impacted long term incidence rates in older adults. These findings suggest a stronger case for the use of the BCG vaccine in school-age populations; in areas with an equivalent level of TB transmission to England; and where the effectiveness of the BCG vaccine has been shown to not reduce with age. They also indicate that a future vaccine, without the reduced effectiveness observed in some geographic areas, may be more effectively targeted at those at school-age that at neonates. However, this depends on the potential duration of protection inferred by vaccination. In addition, they highlight the trade-off between neonatal and school-age vaccination with school-age vaccination reducing overall TB incidence but neonatal vaccination reducing incidence in young children where more severe outcomes are more common. The findings from this thesis may be of use to policy-makers to inform vaccine usage both in the UK and globally. As a by-product of the work conducted in this thesis several open source tools have been developed. These tools maybe used as learning resources, for public engangement, and as part of other research projects.
